### 2017-01-26 manuscript.rmd
added function for adding noise to calibration dates

	# These clade dates are from the KMRR code, they got them from http://timetree.org/
	calibration_times = data.frame(
		age.min =
			c( 92, 167, 8, 296, 29, 162 ), 
		clade = 
			c( "Euarchontoglires", "Mammalia", "Homininae", 
					"Amniota", "Catarrhini", "Theria"
			),
		stringsAsFactors=FALSE
	)
```

@@

@@

		tree = try( 
			ape::chronos( nhx@phylo, calibration=calibration ) 
		)

@@

	#' Calibrate a list of gene trees given a set of speciation node dates
	#' 
	#' @param gene_trees_sub A list of phylogenetic trees as treeio::treedata objects 
	#' with associated character data
	#' @param calibration_times A dataframe with two columns: age.min is the age 
	#' of a clade, and clade is the name of the clade
	#' @param ... Any additional arguments to pass to ape::chronos()
	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
	#' shorter than gene_trees_sub if some calibrations fail.
	#' @export
	calibrate_trees = function( gene_trees_sub, calibration_times, ... ) {
		# Make the trees ultrametric and calibrate the speciation nodes to 
		# specified times
		gene_trees_calibrated = 
			mclapply( 
				gene_trees_sub,
				calibrate_tree, 
				calibration_times=calibration_times,
				...,
				mc.cores=cores
			)

@@

	# Remove tips from the trees that do not have expression data
	gene_trees_sub = mclapply( 
		gene_trees_annotated, 
		drop_empty_tips, 
		min_genes_with_expression=min_genes_with_expression,
		mc.cores=cores 
	)

	# Remove trees with less than min_genes_with_expression tips with expression data
	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]

@@

	# Time calibrate all speciation nodes to the same years
	gene_trees_calibrated = calibrate_trees( gene_trees_sub, calibration_times, model="correlated" )

@@

When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_sub)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau. They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

@@

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. As described above, this is because the oldest speciation event is by definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.

@@

Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making pairwise comparisons of ortholog expression through development between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species. This list of shared genes was different from species pair to species pair. They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

>>>>

	# These clade dates are from the KMRR code, they got them from http://timetree.org/
	calibration_times = data.frame(
		age.min =
			c(  8, 29, 92, 162, 167, 296 ), 
		clade = 
			c( "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" ),
		stringsAsFactors=FALSE
	)
```

@@

	#' Adds some random noise from a normal distribution to calibration
	#' times
	#' 
	#' @param calibration_times A dataframe with two columns: age.min is the age 
	#' of a clade, and clade is the name of the clade
	#' @param sd_fraction this is multiplied by each value to set sd of that value 
	#' @return A dataframe as above, but with noise
	#' @export
	add_noise_calibration_times = function( calibration_times, sd_fraction ) {
		age = calibration_times$age.min
		new_age = NA

		while( NA %in% new_age ) {
			x = sapply( 
				age,
				function( i ){
					rnorm( 1, mean = i, sd = ( sd_fraction * i ) )
				}
			)

			if( ! is.unsorted( x ) ){
				new_age = x
			}

		calibration_times$age.min = new_age
		return( calibration_times )
	}

@@

		tree = try( 
			ape::chronos( nhx@phylo, calibration=calibration, ... ) 
		)

@@

	#' Calibrate a list of gene trees given a set of speciation node dates
	#' 
	#' @param gene_trees_pruned A list of phylogenetic trees as treeio::treedata objects 
	#' with associated character data
	#' @param calibration_times A dataframe with two columns: age.min is the age 
	#' of a clade, and clade is the name of the clade
	#' @param ... Any additional arguments to pass to ape::chronos()
	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
	#' shorter than gene_trees_pruned if some calibrations fail.
	#' @export
	calibrate_trees = function( gene_trees_pruned, calibration_times, ... ) {
		# Make the trees ultrametric and calibrate the speciation nodes to 
		# specified times
		gene_trees_calibrated = 
			mclapply( 
				gene_trees_pruned,
				calibrate_tree, 
				calibration_times=calibration_times,
				...,
				mc.cores=cores
			)

@@

	# Remove tips from the trees that do not have expression data
	gene_trees_pruned = mclapply( 
		gene_trees_annotated, 
		drop_empty_tips, 
		min_genes_with_expression=min_genes_with_expression,
		mc.cores=cores 
	)

	# Remove trees with less than min_genes_with_expression tips with expression data
	gene_trees_pruned = gene_trees_pruned[ ! is.na( gene_trees_pruned ) ]

@@

	# Time calibrate all speciation nodes to the same years
	gene_trees_calibrated = calibrate_trees( gene_trees_pruned, calibration_times, model="correlated" )

@@

When we reanalyze the data with phylogenetic independent contrasts (Fig. 1B), we did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). Of the `r length(gene_trees)` trees that were parsed from ENSMBL Compara [@Herrero:2016jj], `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated pruned trees were used to calculate phylogenetic independent contrasts for Tau. They contain `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

@@

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` years. As described above, this is because the oldest speciation event is be definition the most recent common ancestor of the the species included in the study, but many gene families underwent duplication before this time.

@@

Levin *et al.* [@Levin:2016bp] arrived at this conclusion my making pairwise comparisons of ortholog expression through development between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species. This list of shared genes was different from species pair to species pair. They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.
